Equivalence Testing for Psychological Science

### 2017-11-08 manuscript.tex
Wrote short text box on TOST in R for manuscript

Psychologists should be able to falsify predictions. A common prediction
in psychological research is that an effect exists in the population
that differs from zero. For example, we might predict that priming
American Asian women with their Asian identify will increase their
performence on a math test compared to women in the control condition
who are primed with their female identity. To be able to falsify this
hypothesies, and design a study that allows for strong inferences
({\textbf{???}}), it is important to specify which test result would
\emph{disprove} the hypothesis.

An equivalence test can be used to test whether the observed effect is
surprisingly close to zero, assuming a meaningful effect exists in the
population. The test is a simple variation on the widely used null
hypothesis significance tests, with the small difference that instead of
testing the observed effect against zero, it is tested against a
smallest effect size of interest (SESOI). For example, if a difference
on a math test of 5\% or more is considered meaningful, one could test
against a SESOI of 5\%. If the observed effect is suprisingly closer to
0, assuming a true effect of 5\% or larger existed, one can declare
statistical equivalence, or reject the presence of effects that are
large enough to be considered meaningful. An equivalence test can be
used to falsify hypotheses.

One can conclude statistical equivalence whenever a 90\% confidence
interval around the effect size estimate does not contain the SESOI, or
when two one-sided tests (i.e., examining whether the difference on the
math test is more than -5\%, but less than 5\%) are statistically
significant. By combining null-hypothesis tests and equivalence tests,
four combinations of results can be observed (see Figure 1). Two
outcomes are straightforward to interpret, namely when the effect is
equivalent to zero, and not different from zero (an effect that is not
meaningful), and when the effect is not equivalent, and different from
zero (a meaningful effect). In addition, the effect can be different
from zero and too small to be considered meaningful (an effect which is
not meaningful) or the effect can neither differ from zero, not be
equivalent to zero (an undetermined result, where more data is needed).

Even though equivalence tests are just a small variation of traditional
Frequentist tests (testing against the SESOI, instead of 0), their use
was limited until the availability of user-friendly software to perform
the calculations (Lakens, 2017). In this article, we discuss different
approaches to determining the SESOI for psychological research, and
provide detailed reproducible examples of how to perform power analyses
when designing equivalence tests, and statistical re-analyses of
published psychology experiments.

@@

Equivalence tests are performed against a value that is considered the
smallest effect size of interest (SESOI). The SESOI can sometimes based
on just noticeable differences, which can be objectively determined.
Most often, however, it is a subjective decision that varies across
individuals and time. Both these approaches are valid, and even when a
just noticeable difference can be objectively determined, researchers
might choose to set the SESOI to a larger value. In many research areas,
the SESOI is best based on a cost-benefit analysis. Since both costs and
benefits are necessarily relative, the SESOI will depend on the
researcher who designs the study. The goal of setting a SESOI is to
clearly justify why designing a study that has a high probability of
rejecting effects larger than a specified value contributes to our
knowledge base. Researchers should not aim to determine a SESOI that is
universally valid. The goal is to set a SESOI such that inferences based
on it answer a meaningful question.

@@

An objectively determined SESOI should be based on quantifiable
theoretical predictions, such as computational models. Sometimes, the
only theoretical prediction is that an effect should be noticeable. In
such circumstances, the SESOI can be set based on just noticeable
differences. For example, Burriss and colleagues ({\textbf{???}})
examined whether women displayed an increase in redness in the face
during the fertile phase of their ovulatory cycle. The hypothesis was
that a slighly redder skin signals greater attractiveness and physical
health, and sending this signal to men yields an evolutionary advantage.
This hypothesis requires that the increase in redness can be detected
with the naked eye by men. They collected data from 22 women and showed
that there was indeed an increase in redness of the facial skin of woman
during their fertile period. However, this increase was not large enough
to be noticeable with the naken eye by men, thus falsifying their
hypothesis. Because the just noticeable difference in redness of the
skin can be measured, it is possible to objectively establish the SESOI.

@@

Second, researchers can determine the SESOI based on the literature.
Ideally, researchers would specify the SESOI in their research, but this
is not yet common practice. It is thus up to researchers who build on
earlier work to decide which effect size is too small to be meaningful,
given an earlier study. Simonsohn (2015) recently proposed to set the
SESOI to 33\% of the effect size an earlier study could detect based on
the sample size. For example, consider a study where 100 participants
answered a question, which was analyzed with an one-sample t-test. For a
two-sided test with an alpha of .05 this test had 33\% power to detect a
d = 0.15.

Another justifiable choice we would like to propose here is to use the
smallest observed effect size that could have been statistically
significant in the original study as the SESOI in a replication study.
Based only on the alpha level and the sample size, we can calculate the
criticial test value (e.g., \emph{t}, \emph{F}, \emph{Z}). This critical
test value can also be transformed to a standardized effect size (e.g.,
\(d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }\)),
which can thus be interpreted as a \emph{critical effect size}. All
observed effect sizes smaller than the critical effect size would not be
statistically significant in an original study, given the alpha and
sample size. By setting the SESOI to the critical effect size, an
equivalence test can reject all observed effect sizes that could have
been detected in an earlier study.

Third, researchers can set the SESOI based on the resources they have
available. The amount of data you can collect limits the inferences you
can make. Given a an alpha level and a sample size, researchers can
calculate the smallest effect size that can be rejected with the desired
power in an equivalence test. For example, a researcher who plans to
perform a two-sided one-sample t-test using an alpha of 5\%, based on
data from 100 observations, has 90\% power declare equivalence when
testing a SESOI of d = 0.33. Whether or not a test against a SESOI based
on the available resources contributes to the scientific literature
depends. One should not expect a test based on 11 observations, which
provides 90\% power to reject effects larger than d = 1, is particularly
informative, given that most effects in psychology are substantially
smaller than d = 1. But by transparently reporting the effects one can
detect and reject, based on the study design, researchers can
communicate the information their study contributes.

\[t _ { L } = \frac { M _ { 1} - M _ { 2} - \Delta _ { L } } { \sigma \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } } }, t _ { U } = \frac { M _ { 1} - M _ { 2} - \Delta _ { U } } { \sigma \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } } }
\]

\subsection{Examples}\label{examples}

\section{Equivalence test for
Meta-analysis}\label{equivalence-test-for-meta-analysis}

Hyde, Lindberg, Linn, Ellis, and Williams (2008) report that effect
sizes for gender differences in mathematics tests across the 7 million
students in the US represent trivial differences, where a trivial
difference is specified as an effect size smaller then d = 0.1. The
present a table with Cohen's d and se is reproduced below:

\begin{longtable}[]{@{}ll@{}}
\toprule
Grades & d + se\tabularnewline
\midrule
\endhead
Grade 2 & 0.06 +/- 0.003\tabularnewline
Grade 3 & 0.04 +/- 0.002\tabularnewline
Grade 4 & -0.01 +/- 0.002\tabularnewline
Grade 5 & -0.01 +/- 0.002\tabularnewline
Grade 6 & -0.01 +/- 0.002\tabularnewline
Grade 7 & -0.02 +/- 0.002\tabularnewline
Grade 8 & -0.02 +/- 0.002\tabularnewline
Grade 9 & -0.01 +/- 0.003\tabularnewline
Grade 10 & 0.04 +/- 0.003\tabularnewline
Grade 11 & 0.06 +/- 0.003\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.05 the meta-analysis was significant, Z = 20, p = 2.753624e-89
## Using alpha = 0.05 the equivalence test was significant, Z = -13.33333, p = 7.406413e-41TOST results:
##   Z-value 1 p-value 1 Z-value 2    p-value 2
## 1  53.33333         0 -13.33333 7.406413e-41
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 90% CI Upper Limit 90% CI
## 1         0.05506544         0.06493456
\end{verbatim}

We see that indeed, all effect size estimates are measured with such
high precision, we can conclude that they fall within the equivalence
bound of d = -0.1 and d = 0.1. However, note that all of the effects are
also statistically significant - so, the the effects are statistically
different from zero, and practically equivalent.

\emph{Hyde, J. S., Lindberg, S. M., Linn, M. C., Ellis, A. B., \&
Williams, C. C. (2008). Gender similarities characterize math
performance. Science, 321(5888), 494-495.}

~

~

\section{Independent Groups Equivalence
Test}\label{independent-groups-equivalence-test}

~

\subsection{Independent Groups Student's Equivalence
Test}\label{independent-groups-students-equivalence-test}

Eskine (2013) showed that participants who had been exposed to organic
food were substantially harsher in their moral judgments relative to
those exposed to control (d = 0.81, 95\% CI: {[}0.19, 1.45{]}). A
replication by Moery \& Calin-Jageman (2016, Study 2) did not observe a
significant effect (Control: n = 95, M = 5.25, SD = 0.95, Organic Food:
n = 89, M = 5.22, SD = 0.83). Following Simonsohn's (2015)
recommendation the equivalence bound was set to the effect size the
original study had 33\% power to detect (with n = 21 in each condition,
this means the equivalence bound is d = 0.48, which equals a difference
of 0.384 on a 7-point scale given the sample sizes and a pooled standard
deviation of 0.894). Using a TOST equivalence test with alpha = 0.05,
assuming equal variances, and equivalence bounds of d = -0.43 and d =
0.43 is significant, t(182) = -2.69, p = 0.004. We can reject effects
larger than d = 0.43.

@@

\begin{verbatim}
## Using alpha = 0.05 Student's t-test was non-significant, t(182) = 0.2274761, p = 0.8203089
## Using alpha = 0.05 the equivalence test based on Student's t-test was significant, t(182) = -2.684218, p = 0.003970479TOST results:
##   t-value 1    p-value 1 t-value 2   p-value 2  df
## 1   3.13917 0.0009885653 -2.684218 0.003970479 182
## 
## Equivalence bounds (raw scores):
##   low bound raw high bound raw
## 1        -0.384          0.384
## 
## TOST confidence interval:
##   Lower Limit 90% CI raw Upper Limit 90% CI raw
## 1             -0.1880364              0.2480364
\end{verbatim}

\emph{Moery, E., \& Calin-Jageman, R. J. (2016). Direct and Conceptual
Replications of Eskine (2013): Organic Food Exposure Has Little to No
Effect on Moral Judgments and Prosocial Behavior. Social Psychological
and Personality Science, 7(4), 312-319.
\url{https://doi.org/10.1177/1948550616639649} }

~

\subsection{Independent Groups Welch's Equivalence
Test}\label{independent-groups-welchs-equivalence-test}

Deary, Thorpe, Wilson, Starr, and Whally (2003) report the IQ scores of
79,376 children in Scotland (39,343 girls and 40,033 boys). The IQ score
for girls (M = 100.64, SD = 14.1) girls and boys (100.48, SD = 14.9) was
non-significant. With such a huge sample, we can examine whether the
data is equivalent using very small equivalence bounds, such as -0.05
and 0.05. Because sample sizes are unequal, and variances are unequal,
Welch's t-test is used, and we can conclude IQ scores are indeed
equivalent, based on equivalence bounds of -0.05 and 0.05.

@@

\begin{verbatim}
## Using alpha = 0.05 Welch's t-test was non-significant, t(79260.94) = 1.554136, p = 0.1201559
## Using alpha = 0.05 the equivalence test based on Welch's t-test  was significant, t(79260.94) = -5.490723, p = 2.007585e-08TOST results:
##   t-value 1    p-value 1 t-value 2    p-value 2       df
## 1  8.598995 4.092652e-18 -5.490723 2.007585e-08 79260.94
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1       -0.05         0.05
## 
## Equivalence bounds (raw scores):
##   low bound raw high bound raw
## 1    -0.7252758      0.7252758
## 
## TOST confidence interval:
##   Lower Limit 90% CI raw Upper Limit 90% CI raw
## 1           -0.009341433              0.3293414
\end{verbatim}

\emph{Deary, I. J., Thorpe, G., Wilson, V., Starr, J. M., \& Whalley, L.
J. (2003). Population sex differences in IQ at age 11: The Scottish
mental survey 1932. Intelligence, 31(6), 533-542.}

~

~

\section{One-Sample Equivalence Test}\label{one-sample-equivalence-test}

~

Lakens, Semin, \& Foroni (2012) examined whether the color of Chinese
ideographs (white vs.~black) would bias whether participants judged that
the ideograph correctly translated positive, neutral, or negative words.
In Study 4 the prediction was that participants would judge negative
words to be correctly translated by black ideographs above guessing
average, but no effects were predicted for translations in the other 5
between subject conditions in the 2 (ideograph color: white vs.~black) x
3 (word meaning: positive, neutral, negative) design. The table below is
a summary of the data in all 6 conditions:

\begin{longtable}[]{@{}lllllllll@{}}
\toprule
Color & Valence & M & \% & SD & t & df & p & d\tabularnewline
\midrule
\endhead
White & Positive & 6.05 & 50.42 & 1.50 & 0.15 & 20 & .89 &
.03\tabularnewline
White & Negative & 6.68 & 55.67 & 1.70 & 1.75 & 18 & .10 &
.40\tabularnewline
White & Neutral & 5.95 & 49.58 & 1.40 & 0.87 & 19 & .87 &
.04\tabularnewline
Black & Positive & 6.45 & 53.75 & 1.91 & 1.06 & 19 & .30 &
.24\tabularnewline
Black & Negative & 6.95 & 57.92 & 1.15 & 3.71 & 19 & .00 &
.80\tabularnewline
Black & Neutral & 5.71 & 47.58 & 1.79 & 0.73 & 20 & .47 &
.16\tabularnewline
\bottomrule
\end{longtable}

With 19 to 21 participants in each between subject condition, the study
had 80\% power to detect equivalence with equivalence bounds of -0.68 to
d = 0.68.

\begin{verbatim}
## The required sample size to achieve 80 % power with equivalence bounds of -0.68 and 0.68 is 19
\end{verbatim}

\begin{verbatim}
## [1] 18.52043
\end{verbatim}

When we perform 5 equivalence tests using equivalence bounds of -0.68
and 0.68, using a Holm-Bonferroni sequential procedure to control error
rates, we can conclude statistical equivalence for the data in row 1, 3,
and 6, but the conclusion is undetermined for tests 2 and 4, where the
data is neither significantly different from guessing average, nor
statistically equivalent. For example, performing the test for the data
in row 6:

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.05 the NHST one-sample t-test was non-significant, t(19) = -0.724536, p = 0.4775649
## Using alpha = 0.05 the equivalence test was significant, t(19) = 2.316516, p = 0.01592727TOST results:
##   t-value 1  p-value 1 t-value 2    p-value 2 df
## 1  2.316516 0.01592727 -3.765588 0.0006543016 19
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1       -0.68         0.68
## 
## Equivalence bounds (raw scores):
##   low bound raw high bound raw
## 1       -1.2172         1.2172
## 
## TOST confidence interval:
##   Lower Limit 90% CI raw Upper Limit 90% CI raw
## 1             -0.9820961              0.4020961
\end{verbatim}

It is clear I should have been more tentative in my conclusions in this
study. Not only can I not conclude equivalence in some of the
conditions, the equivalence bound I had 80\% power to detect is very
large, meaning the possibility that there are theoretically interesting
but smaller effects remains.

\emph{Lakens, D., Semin, G. R., \& Foroni, F. (2012). But for the bad,
there would not be good: Grounding valence in brightness through shared
relational structures. Journal of Experimental Psychology: General,
141(3), 584-594. \url{https://doi.org/10.1037/a0026468} }

~

~

\section{Equivalence test for
correlations}\label{equivalence-test-for-correlations}

~

Olson, Fazio, and Hermann (2007) reported correlations between implicit
and explicit measures of self-esteem, such as the IAT, Rosenberg's
self-esteem scale, a feeling thermometer, and trait ratings. In Study 1
71 participants completed the self-esteem measures. Because no
equivalence bounds are mentioned, we can see which equivalence bounds
the researchers would have 80\% power to detect.

\begin{verbatim}
## The required sample size to achieve 80 % power with equivalence bounds of -0.24 and 0.24 is 71 pairs
\end{verbatim}

\begin{verbatim}
## [1] 70.05703
\end{verbatim}

With 71 pairs of observations between measures, the researchers have
80\% power for equivalence bounds of r = -0.24 and r = 0.24.

The correlations observed by Olson et al (2007), Study 1, are presented
in the table below (significant correlations are flagged by an
asteriks).

\begin{longtable}[]{@{}lllll@{}}
\toprule
Measure & IAT & Rosenberg & Feeling thermometer & Trait
ratings\tabularnewline
\midrule
\endhead
IAT & - & -.12 & -.09 & -.06\tabularnewline
Rosenberg & & - & .62* & .09\tabularnewline
Feeling thermometer & & & - & .29*\tabularnewline
Trait ratings & & & & -\tabularnewline
\bottomrule
\end{longtable}

We can test each correlation, for example the correlation between the
IAT and the Rosenberg self-esteem scale of -0.12, given 71 participants,
and using equivalence bounds of r\_U = 0.24 and r\_L = -0.24.

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.05 the NHST t-test was non-significant, p = 0.6191585
## Using alpha = 0.05 the equivalence test was non-significant, p = 0.06386793TOST results:
##     p-value 1  p-value 2
## 1 0.005971455 0.06386793
## 
## Equivalence bounds (r):
##   low bound r high bound r
## 1       -0.24         0.24
## 
## TOST confidence interval:
##   Lower Limit 90% CI raw Upper Limit 90% CI raw
## 1             -0.2538652              0.1384997
\end{verbatim}

We see that none of the correlations can be declared to be statistically
equivalent. Instead, the tests yield undetermined outcomes: the
correlations are not significantly different from 0, nor statistically
equivalent.

>>>>

Psychologists should be able to falsify predictions. A common prediction
in psychological research is that an effect exists in the population
that differs from zero. For example, we might predict that priming
American Asian women with their Asian identity will increase their
performance on a math test compared to women who are primed with their
female identity. To be able to falsify this hypothesis, and design a
study that allows for strong inferences ({\textbf{???}}), it is
important to specify which test result would \emph{disprove} the
hypothesis.

An equivalence test can be used to test whether the observed effect is
surprisingly small, assuming a meaningful effect exists in the
population. The test is a simple variation of the widely used null
hypothesis significance tests (NHST). To understand the idea behind
equivalence tests, it is useful to realize the null hypothesis we test
against can be any numerical value. When we compare two groups, we often
test whether the difference between these groups is zero, but we may
sometimes want to test against other values than zero. Imagine a
researcher who is interested in voluntary participation in a national
program to train young infants' motor skills. The researcher wants to
test whether more boys than girls are brought into the program by their
parents. One could test whether the difference between boys and girls is
zero. However, this would ignore the fact that the human sex ratio of
newborn boys to girls is not exactly 1:1, and we should not expect 50\%
of participants to be boys. On average, 103 boys are born for every 100
girls (CIA, 2017), so approximately 50.74\% of applicants should be
boys, and 49.26\% should be girls (with 0.5074/0.4926≈1.03). If boys and
girls were exactly equally likely to be brought into the program by
their parents, we would not expect a difference of zero, but 1.5\% more
boys. This can be used as a null hypothesis: Rather than testing against
a difference in proportions of 0, the researcher tests against 0.015.

Even though 0.015 might be a more sensible value for a null hypothesis,
our researcher knows that the true sex ratio in their population will
differ slightly from that in the entire world. Rather than testing a
point null hypothesis (setting \(H_0\) to one specific value), the
researcher decides to define a range of values around the difference in
proportions of 0.015 that can be considered trivially small, even when
the true ratio in the population is not exactly 0, or even 0.015. The
researcher can for example test if the difference is smaller than
-0.005, or larger than 0.035. This test against two bounds, with \(H_0\)
as a range rather than one value, is known as a \emph{minimal effects
test} (Murphy, Myors, \& Wolach, 2014).(Murphy et al. 2014)

However, even if the researcher would consider any difference in
percentages smaller than \(-0.5 \%\) or larger than \(3.5 \%\) a
rejection of the null hypothesis, not every difference is large enough
to matter in practice. As long as gender differences in the
participation rate are not too extreme, there is no need to spend money
to address the gender imbalance in the voluntary participation rate, and
all money can be spent on the core training program. To test whether the
gender difference in participants is not large enough to matter, the
researcher can perform an \emph{equivalence test}. Equivalence tests
examine whether the presence of gender differences that are large enough
to matter can be rejected. After an extensive discussion with experts,
the researcher decides that as long as the difference in proportions is
not larger than 6\% the researcher can act as if the gender difference
is too small to intervene. Given an expected true difference in the
population of 0.015, the researcher will tests if the observed
difference in the data is smaller than 0.075, and larger than -0.055. If
differences more extreme than both these boundary values can be rejected
in two one-sided tests, the researcher will conclude statistical
equivalence, and no money will be spent on addressing a gender
difference in participation.

One can conclude statistical equivalence whenever a 90\% confidence
interval around the observed estimate does not contain the smallest
effect size of interest (SESOI). Figure 1 illustrates the possible
conclusions we can draw from null hypothesis, minimal effect, and
equivalence tests.

Even though equivalence tests are just a small variation of traditional
frequentist null-hypothesis significance tests (and test against the
SESOI, instead of a value of 0), their use was limited until the
availability of user-friendly software to perform the calculations
(Lakens, 2017). In this article, we provide several examples of
equivalence tests which illustrate the procedure. Furthermore, we
discuss different approaches to determining the SESOI for psychological
research, and provide detailed reproducible examples of how to perform
power analyses when designing equivalence tests, and statistical
re-analyses of published psychology experiments.

@@

Equivalence tests are performed against a value that is considered the
smallest effect size of interest (SESOI). The SESOI can sometimes be
based on just noticeable differences, which can be objectively
determined. Most often, however, it is a subjective decision that varies
across individuals and time. The SESOI is ideally based on a
cost-benefit analysis. Since both costs and benefits are necessarily
subjective, the SESOI will depend on the researcher who designs the
study. The goal of setting a SESOI is to clearly justify why designing a
study that has a high probability of rejecting effects larger than a
specified value contributes to our knowledge base. A SESOI should be
chosen such that inferences based on it answer a meaningful question.

@@

An objectively determined SESOI should be based on quantifiable
theoretical predictions, such as computational models. Sometimes, the
only theoretical prediction is that an effect should be noticeable. In
such circumstances, the SESOI can be set based on just noticeable
differences. For example, Burriss and colleagues ({\textbf{???}})
examined whether women displayed an increase in redness in the face
during the fertile phase of their ovulatory cycle. The hypothesis was
that a slightly redder skin signals greater attractiveness and physical
health, and sending this signal to men yields an evolutionary advantage.
This hypothesis requires that the increase in redness can be detected
with the naked eye by men. They collected data from 22 women and showed
that there was indeed an increase in redness of the facial skin of woman
during their fertile period. However, this increase was not large enough
to be noticeable with the naked eye by men, thus falsifying their
hypothesis. Because the just noticeable difference in redness of the
skin can be measured, it is possible to objectively establish the SESOI.

@@

Second, researchers can determine the SESOI based on previous work in
the literature. Ideally, researchers would specify their SESOI in
studies they report, but this is not yet common practice. It is thus up
to researchers who build on earlier work to decide which effect size is
too small to be meaningful, given an earlier study. Simonsohn (2015)
recently proposed to set the SESOI to the effect size that an earlier
study on the same effect would have 33\% power to detect based on the
sample size. This is called the \enquote{small telescopes} approach. For
example, consider a study where 100 participants answered a question,
which was analyzed with an one-sample t-test. For a two-sided test with
an alpha of 0.05 this test had 33\% power to detect a d = 0.15.

Another justifiable choice for the SESOI we would like to propose here
is to use the smallest observed effect size that could have been
statistically significant in the original study. In other words, we
decide that effects that could not have yielded a \emph{p} \textless{}
0.05 in the original study (if this is your preferred alpha threshold)
will not be considered meaningful in the replication either, even if
effects of this size are statistically significant in the replication
study. Based only on the alpha level and the sample size, we can
calculate the critical test value (e.g., \emph{t}, \emph{F}, \emph{Z}).
This critical test value can also be transformed to a standardized
effect size (e.g.,
\(d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }\)),
which can thus be interpreted as a \emph{critical effect size}. All
observed effect sizes smaller than the critical effect size would not be
statistically significant in an original study, given the alpha and
sample size of that study. By setting the SESOI to the critical effect
size, an equivalence test can reject all observed effect sizes that
could have been detected in an earlier study.

Third, researchers can set the SESOI based on the resources they have
available. The amount of data you can collect limits the inferences you
can make. Given a an alpha level and a sample size, researchers can
calculate the smallest effect size that one has sufficient power to
detect. This approach is suggested here to provide researchers with a
way to think about setting the SESOI for a study in a line of research
where no quantifiable predictions are made, and where no previous
studies have been performed. Observing statistical equivalence for a
SESOI based on this approach allows one to reject effects that can be
reliably detected with the given sample size. For example, a researcher
who plans to perform a two-sided one-sample \emph{t}-test using an alpha
of 5\%, based on data from 100 observations, has 90\% power to detect an
effect of d = 0.364. Using equivalence bounds of d = -0.364 and d =
0.364 in an equivalence test allows one to reject effects the study had
high statistical power to detect. In this case, concluding equivalence
would suggest one can reject effects that can reasonably be detected
with N = 100. Whether or not this is interesting depends in part on how
typical the chosen sample size is for a research area. One should not
expect an equivalence test based on 15 observations testing against a
SESOI of d = -1 and d = 1 to substantially increase our knowledge, given
that most effects in psychology are substantially smaller than these
bounds. But by transparently reporting the effects one can detect and
reject, based on the study design, researchers can communicate the
information their study contributes. When there are no previous studies
to build on, and no theoretical predictions, specifying a SESOI based on
the alpha level and a sample size can provide a starting point for a
discussion on what a reasonable SESOI is. Nevertheless, the value of
setting a SESOI based on this procedure is limited, and the equivalence
test does not reject values that are theoretically interesting.

\subsection{Example 1: Statistically Equivalent and Not Statistically
Different}\label{example-1-statistically-equivalent-and-not-statistically-different}

Banerjee, Chatterjee, \& Sinha (2012) reported that participants who had
been asked to describe an unethical deed from their past judged the room
to be darker than participants who had been asked to describe an ethical
deed (\(M_{unethical}= 4.71\), \({SD}_{unethical}= 0.85\),
\(M_{ethical}=5.30\), \({SD}_{ethical}=0.97\), \(t(38)= 2.03\),
\(p= 0.05\), \(d= 0.65\)). A close replication by Brandt, IJzerman, \&
Blanken (2014) found no significant effect (\(t(98)=0.56\), \(p=0.57\),
\(d=0.11\)). The smallest effect size that could have been statistically
significant in the original study is
\(d_{crit}= 2.02 \sqrt { \frac { 1} {20} + \frac { 1} {20} }=0.64\).
Using this as our SESOI for a TOST with Welch's \emph{t}-test for
independent samples --- resulting in equivalence bounds of
\({\Delta}_L=-0.64\) and \({\Delta}_U=0.64\) --- we indeed find that the
effect reported by the replication study is statistically equivalent,
\(t(97.78)=-2.64\), \(p=0.00\).

@@

\begin{verbatim}
## Using alpha = 0.05 Welch's t-test was non-significant, t(97.78283) = 0.5649507, p = 0.5734012
## Using alpha = 0.05 the equivalence test based on Welch's t-test  was significant, t(97.78283) = -2.638058, p = 0.004851125TOST results:
##   t-value 1    p-value 1 t-value 2   p-value 2       df
## 1   3.76796 0.0001407233 -2.638058 0.004851125 97.78283
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1  -0.6401696    0.6401696
## 
## Equivalence bounds (raw scores):
##   low bound raw high bound raw
## 1    -0.7302364      0.7302364
## 
## TOST confidence interval:
##   Lower Limit 90% CI raw Upper Limit 90% CI raw
## 1              -0.249788               0.507388
\end{verbatim}

\subsection{Example 2: Statistically Equivalent and Statistically
Different}\label{example-2-statistically-equivalent-and-statistically-different}

@@

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = 20, p = 5.507248e-89
## Using alpha = 0.005 the equivalence test was significant, Z = -13.33333, p = 7.406413e-41TOST results:
##   Z-value 1 p-value 1 Z-value 2    p-value 2
## 1  53.33333         0 -13.33333 7.406413e-41
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1         0.05227251         0.06772749
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-2.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = 20, p = 5.507248e-89
## Using alpha = 0.005 the equivalence test was significant, Z = -30, p = 4.906714e-198TOST results:
##   Z-value 1 p-value 1 Z-value 2     p-value 2
## 1        70         0       -30 4.906714e-198
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1         0.03484834         0.04515166
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-3.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -5, p = 5.733031e-07
## Using alpha = 0.005 the equivalence test was significant, Z = 45, p = 0TOST results:
##   Z-value 1 p-value 1 Z-value 2 p-value 2
## 1        45         0       -55         0
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.01515166       -0.004848341
\end{verbatim}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -5, p = 5.733031e-07
## Using alpha = 0.005 the equivalence test was significant, Z = 45, p = 0TOST results:
##   Z-value 1 p-value 1 Z-value 2 p-value 2
## 1        45         0       -55         0
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.01515166       -0.004848341
\end{verbatim}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -5, p = 5.733031e-07
## Using alpha = 0.005 the equivalence test was significant, Z = 45, p = 0TOST results:
##   Z-value 1 p-value 1 Z-value 2 p-value 2
## 1        45         0       -55         0
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.01515166       -0.004848341
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-4.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -10, p = 1.523971e-23
## Using alpha = 0.005 the equivalence test was significant, Z = 40, p = 0TOST results:
##   Z-value 1 p-value 1 Z-value 2 p-value 2
## 1        40         0       -60         0
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.02515166        -0.01484834
\end{verbatim}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -10, p = 1.523971e-23
## Using alpha = 0.005 the equivalence test was significant, Z = 40, p = 0TOST results:
##   Z-value 1 p-value 1 Z-value 2 p-value 2
## 1        40         0       -60         0
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.02515166        -0.01484834
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-5.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = -3.333333, p = 0.0008581207
## Using alpha = 0.005 the equivalence test was significant, Z = 30, p = 4.906714e-198TOST results:
##   Z-value 1     p-value 1 Z-value 2     p-value 2
## 1        30 4.906714e-198 -36.66667 1.241408e-294
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1        -0.01772749       -0.002272512
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-6.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = 13.33333, p = 1.481283e-40
## Using alpha = 0.005 the equivalence test was significant, Z = -20, p = 2.753624e-89TOST results:
##   Z-value 1 p-value 1 Z-value 2    p-value 2
## 1  46.66667         0       -20 2.753624e-89
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1         0.03227251         0.04772749
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-7.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Using alpha = 0.005 the meta-analysis was significant, Z = 20, p = 5.507248e-89
## Using alpha = 0.005 the equivalence test was significant, Z = -13.33333, p = 7.406413e-41TOST results:
##   Z-value 1 p-value 1 Z-value 2    p-value 2
## 1  53.33333         0 -13.33333 7.406413e-41
## 
## Equivalence bounds (Cohen's d):
##   low bound d high bound d
## 1        -0.1          0.1
## 
## TOST confidence interval:
##   Lower Limit 99% CI Upper Limit 99% CI
## 1         0.05227251         0.06772749
\end{verbatim}
